---
title: "data_analysis_LSA"
format: html
editor: visual
---

## Notes

------------------------------------------------------------------------

## Libraries

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(forcats)
library(lme4)
library(corrplot)
library(clipr)
```

## Import data

import data

```{r}
raw_df <- read.csv("../data/pilot-responses-combined.csv")
raw_df <- raw_df %>% 
  group_by(subject_id) %>%
  mutate(ai_score = sum(cognitive_1, cognitive_2, 
                         affective_1, affective_2, 
                         behavioral_1, behavioral_2)/6)
```

```{r}
head(raw_df)
```

## Add Response Categories

```{r}
categories <- list(
  content     = c("rainbow", "horizon", "rain", "light", "sunlight", "colors", "prism", "the", "color",
                    "informative", "descriptive", "sun", "when", "arc", "arch", "beyond",
                    "clouds", "colours", "earth", "ending", "explainer", "explanation",
                    "forms", "freedom", "god", "green", "hits", "peace", "promise",
                    "rainbaw", "rainbows", "red", "reflecting", "shape", "single", "sky",
                    "spectrum", "storylike", "sunshine", "two ends", "vision", "water",
                    "weather", "wordy"),
  gender      = c("woman", "female", "male", "feminine", "girl", "man", "dude", "girly", "lady",
                    "masculine"),
  age         = c("older", "young", "elderly", "old"),
  affect      = c("calm", "enthusiastic", "calming", "curious", "passionate", "angry", "attentive",
                    "disinterested", "energetic", "enthused", "euphoric", "excited",
                    "happiness", "happy", "invested", "joyful", "moody", "nonchalant",
                    "peaceful", "pensive", "relaxed", "stoked"),
  personality = c("confident", "gentle", "kind", "assertive", "caring", "cautious", "dominant", "entitled",
                    "forceful", "friendly", "opinionated", "outgoing", "patient", "sassy",
                    "snooty", "stern", "stoic", "sweet"),
  competence  = c("smart", "educated", "science", "professional", "scientific", "wise", "academic",
                    "accurate", "articulate", "credible", "detailed", "idiot", "impaired",
                    "important", "informed", "intelligent", "knowledgeable", "percise",
                    "profound", "proper", "stupid", "thorough", "thoughtful",
                    "unintelligent", "unprofessional", "well spoken"),
  naturalness = c("fake", "artificial", "ai generated", "digital", "inauthentic", "manufactured", "natural",
                    "robotic", "scripted", "unrealistic"),
  prosody     = c("cadence", "emphasis", "even", "flat", "monotone", "monotoned", 
                  "monotonous", "powerful", "questioning", "slow"),
  voice_quality = c("clear", "deep", "raspy", "grainy", "nasal", "nasally", "rich", "full", "smooth", "whine"), 
  intensity = c("loud", "projective", "quiet", "soft-spoken", "softspoken"), 
  persona     = c("storyteller", "surfer", "teacher", "cartoon", "documentary", "educator", "guider",
                    "instructor", "leprechaun", "morgan freeman", "preppy",
                    "school", "scientist", "smoker", "story teller", "student", "tv voice",
                    "valley girl", "weed", "wizard", "grandma"),
  appearance    = c("black", "white",  "blonde"),
  locale = c("american", "california", "californian", "new yorker", "city", "culture", "urban"),
  evaluation  = c("annoying", "beautiful", "interesting", "boring", "confusing", "cool", "corny",
                    "creative", "dreamy", "forgettable", "fun", "good", "grating", "odd",
                    "pleasant", "quality", "tiring", "dramatic", "vivid", "colorful", "emotive", "exaggerated"))

assign_category <- function(response) {
  if (is.na(response) || response == "") return(NA_character_)
  for (cat_name in names(categories)) {
    if (response %in% categories[[cat_name]]) {
      return(cat_name)
    }
  }
  return("unknown")
}

analysis_df <- raw_df %>%
  mutate(across(
    .cols = c(word_1, word_2, word_3, word_4, word_5),
    .fns  = ~ sapply(tolower(trimws(.x)), assign_category),
    .names = "category_{.col}"
  )) %>%
  rename_with(~ gsub("category_word_", "category_", .x), starts_with("category_word_"))

# add broad characterizations
social_cats   <- c("gender","age","personality","competence","persona",
                   "appearance","locale","affect")
acoustic_cats <- c("voice_quality","prosody","intensity")

analysis_df <- analysis_df %>%
  rowwise() %>%
  mutate(
    n_social   = sum(c_across(category_1:category_5) %in% social_cats),
    n_acoustic = sum(c_across(category_1:category_5) %in% acoustic_cats),
    n_content  = sum(c_across(category_1:category_5) == "content"),
    any_artificial = any(c_across(word_1:word_5) %in% 
                           c("fake","artificial","ai generated","digital","inauthentic",
                             "manufactured","robotic","scripted","unrealistic"))
  ) %>%
  ungroup()
```

## Data Analysis - Across Speakers

```{r}
ggplot(analysis_df, aes(x = ai_score, 
                        y = humanness_rating)) +
  geom_point(color = "black", 
             size = 3,
             alpha = 0.75) +       
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() 

cor(analysis_df$ai_score, analysis_df$humanness_rating, method = "pearson")
```

Humanness rating is minimally correlated with AI attitude score (Pearson's r = 0.096).

## Data Analysis - By Speaker

```{r}
analysis_df %>% 
  group_by(voice) %>% 
  summarise(
    mean = mean(humanness_rating),
    sd = sd(humanness_rating), 
    median = median(humanness_rating)
  )
```

```{r}
valence_counts_voice <- analysis_df %>%
  select(subject_id, word_1:word_5, voice) %>%
  pivot_longer(word_1:word_5, names_to = "slot", values_to = "word") %>%
  mutate(word = tolower(trimws(word))) %>%
  left_join(valence_lookup, by = "word") %>%
  group_by(voice) %>%
  summarise(
    n_positive  = sum(valence == "positive",  na.rm = TRUE),
    n_negative  = sum(valence == "negative",  na.rm = TRUE),
    n_ambiguous = sum(valence == "ambiguous", na.rm = TRUE),
    .groups = "drop"
  )
valence_counts_voice
```

visualizations

```{r}

ggplot(analysis_df, aes(voice, humanness_rating, fill = voice)) + 
  geom_violin() +
  scale_fill_brewer(palette = "Set3") +
  geom_jitter(width = 0.05) +
  theme_minimal()

```

```{r}
long_df %>% 
  filter(cat != "content") %>% 
  ggplot() +
  geom_bar(aes(y = voice,
               fill = cat),
           position = "fill") +
  coord_flip() +
  labs(x = "Proportion of Non-Content Responses",
       y = "Voice",
       fill = "Category")

ggsave(file="./data_analysis_files/by-voice-categories.png",width=10,height=6)
```

## Sankey prep

```{r}
category_counts <- analysis_df %>%
  pivot_longer(
    cols = category_1:category_5,
    names_to = "word_number",
    values_to = "category"
  ) %>%
  count(category, sort = TRUE)
```

```{r}
View(category_counts)
```

## AI attitude

add a col for averaged AI attitude per listener

```{r}
analysis_df <- analysis_df %>%
  rowwise() %>%
  mutate(ai_attitude_score = rowMeans(across(c(cognitive_1, behavioral_1, affective_1, cognitive_2, affective_2, behavioral_2))))
```

```{r}

ai_attitude_plot <- ggplot(analysis_df, aes(x = ai_attitude_score, y = fct_reorder(source_file, ai_attitude_score))) +
  geom_point(size = 1.5, alpha = 0.7) +
  labs(x = "AI attitude score", y = "participants") + 
  theme_minimal() +
  theme(axis.text.y = element_blank())
```

```{r}
ai_attitude_plot
```

Density + box (i dont really like this)

```{r}
ggplot(analysis_df, aes(x = ai_attitude_score)) +
  geom_histogram(aes(y = after_stat(density)), 
                 fill = "steelblue", 
                 color = "white", 
                 alpha = 0.4,
                 bins = 30) +       
  geom_density(fill = "skyblue", alpha = 0.5) +
  labs(title = "Distribution of AI Attitude Scores", x = "AI Attitude Score", y = "Density") +
  theme_minimal()
```

```{r}
ggplot(analysis_df, aes(x = "", y = ai_attitude_score)) +
  geom_boxplot(fill = "lightpink", alpha = 0.5, width = 0.3) +
  labs(title = "Distribution of AI Attitude Scores", 
       x = "All participants") +
  geom_jitter(width = 0.05)+
  theme_minimal()
```

## Question: Social attributions persist, even at low humanness?

```{r}
long_df <- analysis_df %>%
  pivot_longer(category_1:category_5, names_to="slot", values_to="cat") %>%
  filter(cat != "content") %>%
  mutate(is_social = cat %in% social_cats,
         slot_num = as.integer(gsub("category_","",slot)))

subj_df <- long_df %>% # subject-clustering accounts for between-subj design
  group_by(subject_id, voice, humanness_rating) %>%
  summarise(n_social = sum(is_social), n_total = n(), .groups="drop")

# models for pilot data (random effects result in singularity)
m_social_1 <- glm(cbind(n_social, n_total - n_social) ~ scale(humanness_rating) + voice,
              data = subj_df, family = binomial)
summary(m_social_1)

m_social_2 <- glm(cbind(n_social, n_total - n_social) ~ scale(humanness_rating),
                  data = subj_df, family = binomial)
summary(m_social_2)

# which model better accounts for variance? 
anova(m_social_1, m_social_2, test = "LRT")
```

Model results suggest that perceived humanness modulates the attribution of social descriptors, such that higher "humanness" ratings positively predict a greater rate of social descriptors (β = 0.46, p \< .001).

Adding voice as a fixed effect did not attenuate the humanness slope (β = 0.61, p \< .001), and voice did not improve model fit (LRT p = .21), indicating the effect of humanness is not driven by stimulus-specific variation

Note: Because the design was between subjects, I modeled social attribution at the participant level rather than the descriptor level. For each participant, I computed the number of non-content descriptors that could be described as "social" (gender, age, personality, competence, persona, appearance, locale, affect) our of the total non-content descriptors produced, and then I fit a binomial GLM predicting this proportion from z-scored humanness rating.

```{r}
# predicted probabilities
newdat <- data.frame(humanness_rating = c(0, 30, 50, 66.6, 100))
newdat$z <- (newdat$humanness_rating - mean(analysis_df$humanness_rating)) / 
  sd(analysis_df$humanness_rating)
pred <- predict(m_social_2, newdata = data.frame(humanness_rating = newdat$humanness_rating),
                type = "link", se.fit = TRUE)
newdat$p     <- plogis(pred$fit)
newdat$lo    <- plogis(pred$fit - 1.96 * pred$se.fit)
newdat$hi    <- plogis(pred$fit + 1.96 * pred$se.fit)
newdat

newdat <- data.frame(humanness_rating = seq(0, 100, by = 2))
pred <- predict(m_social_2, newdata = data.frame(humanness_rating = newdat$humanness_rating,
                                             voice = "gracie_valley"),  # arbitrary ref level
                type = "link", se.fit = TRUE)
newdat$p  <- plogis(pred$fit)
newdat$lo <- plogis(pred$fit - 1.96 * pred$se.fit)
newdat$hi <- plogis(pred$fit + 1.96 * pred$se.fit)

# visualization
ggplot() +
  geom_ribbon(data = newdat, aes(humanness_rating, ymin = lo, ymax = hi), 
              alpha = 0.2, fill = "steelblue") +
  geom_line(data = newdat, aes(humanness_rating, p), color = "steelblue", size = 1) +
  geom_jitter(data = subj_df, 
              aes(humanness_rating, n_social / n_total),
              width = 0.5, height = 0.02, alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey") +
  labs(x = "Humanness Rating", 
       y = "P(social or psychological | non-content descriptor)") +
  theme_minimal()
```

Social attribution rate remained substantial across the humanness range, even for voices rated near the bottom of the scale, roughly 37–49% of non-content descriptors were social, well above a floor of zero.

## Question: Transition Probabilities?

```{r}
transitions <- analysis_df %>%
  select(subject_id, voice, category_1:category_5) %>%
  pivot_longer(category_1:category_5, names_to="slot", values_to="cat") %>%
  mutate(slot = as.integer(gsub("category_","",slot))) %>%
  filter(cat != "content") %>%
  arrange(subject_id, slot) %>%
  group_by(subject_id) %>%
  mutate(next_cat = lead(cat)) %>%
  filter(!is.na(next_cat)) %>%
  ungroup()

# raw transition probability matrix
trans_mat <- transitions %>%
  count(cat, next_cat) %>%
  group_by(cat) %>%
  mutate(p = n / sum(n)) %>%
  ungroup()

# add marginal probabilities 
marginals <- transitions %>%
  count(next_cat) %>%
  mutate(p_marginal = n / sum(n)) %>%
  select(next_cat, p_marginal)

# matrix with marginals 
trans_mat_w_margin <- trans_mat %>%
  left_join(marginals, by = "next_cat") %>%
  mutate(lift = p / p_marginal)
```

Unsure what exactly to make of this. There is some same-category momentum (competence -\> competence; persona -\> persona; affect -\> affect; personality -\> personality). This probably reflects some sort of participant response strategy. There also seems to be a broader pattern of acoustics –\> social.

Perhaps these can be framed as predictions about what patterns we will observe in the larger dataset (once collected)?

## Question: AI attitude as predictor?

```{r}
valence_lookup <- tibble(
  word = c(
    "beautiful", "pleasant", "cool", "fun", "dreamy", "vivid", "colorful",
    "annoying", "boring", "grating", "tiring", "forgettable", "odd", "corny", "confusing",
    "interesting", "dramatic", "emotive", "exaggerated"
  ),
  valence = c(
    rep("positive", 7),
    rep("negative", 8),
    rep("ambiguous", 4)
  )
)

# participant valence_counts
valence_counts <- analysis_df %>%
  select(subject_id, word_1:word_5) %>%
  pivot_longer(word_1:word_5, names_to = "slot", values_to = "word") %>%
  mutate(word = tolower(trimws(word))) %>%
  left_join(valence_lookup, by = "word") %>%
  group_by(subject_id) %>%
  summarise(
    n_positive  = sum(valence == "positive",  na.rm = TRUE),
    n_negative  = sum(valence == "negative",  na.rm = TRUE),
    n_ambiguous = sum(valence == "ambiguous", na.rm = TRUE),
    .groups = "drop"
  )

eval_df <- analysis_df %>% left_join(valence_counts, by = "subject_id")

# restrict to subj with at least one valenced evaluation
eval_sub <- eval_df %>%
  filter(n_positive + n_negative > 0) %>%
  mutate(p_positive = n_positive / (n_positive + n_negative))

# mean proportion positive by any_artificial
prop_summary <- eval_sub %>%
  group_by(any_artificial) %>%
  summarise(
    n              = n(),
    mean_p_positive = mean(p_positive),
    sd_p_positive   = sd(p_positive),
    se_p_positive   = sd(p_positive) / sqrt(n),
    .groups = "drop"
  )
prop_summary

# mean ai_score by any_artificial
ai_summary <- eval_sub %>%
  group_by(any_artificial) %>%
  summarise(
    n              = n(),
    mean_ai_score  = mean(ai_score),
    sd_ai_score    = sd(ai_score),
    se_ai_score    = sd(ai_score) / sqrt(n),
    .groups = "drop"
  )
ai_summary

ggplot(eval_sub, aes(x = ai_score, 
                     y = p_positive, 
                     color = any_artificial)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.15) +
  scale_color_brewer(palette = "Set1",
                     name = "Any artificial\ndescriptor") +
  labs(
    x = "AI attitude score (higher = more positive toward AI)",
    y = "Proportion positive valence") +
  theme_minimal()
```

Results inconclusive. Heavily lacking any cases of artificial descriptions paired with valenced evaluations.

```{r}
glm(p_positive ~ scale(humanness_rating)*scale(ai_score), 
    data = eval_sub,
    family = binomial) %>%
  summary()
```

## word cloud prep

(includes content words...)

```{r}
gracie_words <- analysis_df %>% filter(voice == "gracie_valley")
gracie_words <- gracie_words[,c("word_1", "word_2", "word_3", "word_4", "word_5")]
gracie_words_lst <- unlist(gracie_words) %>% as.character()
write_clip(gracie_words_lst)
```

```{r}
vince_words <- analysis_df %>% filter(voice == "vince_duglus")
vince_words <- vince_words[,c("word_1", "word_2", "word_3", "word_4", "word_5")]
vince_words_lst <- unlist(vince_words) %>% as.character()
write_clip(vince_words_lst)
```
